---
title: "Medicare ACO Spending Patterns: Extension to 2014–2024"
author: "Eric Walczyk"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5, dpi = 220)
suppressPackageStartupMessages({
  library(dplyr); library(readr); library(purrr); library(stringr); library(tidyr)
  library(janitor); library(ggplot2); library(fixest); library(modelsummary)
})
setFixest_notes(FALSE)

# Paths are relative to this file (analysis/), so knitting works from a fresh clone
data_dir <- file.path("..", "data")
out_dir  <- file.path("..", "outputs")
fig_dir  <- file.path(out_dir, "figures")
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE)
```

## Overview

The replication (`ACO_Final.rmd`) covers 2014–2017. This file extends the same
models through performance year 2024 using the MSSP Performance Year Financial
and Quality Results PUFs, and tests whether the original findings survive the
2019 *Pathways to Success* redesign and the COVID-19 years.

Differences from the replication:

- **Two-way fixed effects.** All models include ACO *and* performance-period
  fixed effects (`fixest`), with standard errors clustered by ACO. ACOs observed
  in only one period are dropped, since they add no within-ACO variation.
- **Outliers trimmed within period.** Savings rates rise from a median of ~0.5%
  in 2014 to ~4% in 2024, so the 1.5 × IQR rule is applied within each period
  rather than to the pooled sample.
- **Two ACO ID systems.** The 2014–2017 PUFs use a masked `ACO_Num`; 2018+ use
  the CMS `ACO_ID`. There is no official crosswalk, so the main models treat the
  two eras as separate units. The year-trend models and a robustness check
  link them by ACO name.
- **2019 has two files.** `2019.csv` holds full-year and January–June ACOs;
  `2019A.csv` holds ACOs that started new agreements on July 1, 2019. Each is
  its own period.

### Load and Harmonize Data

```{r}
cap_vars <- c("capann_inp_all", "capann_opd", "capann_snf", "capann_pb",
              "capann_ambpay", "capann_hha", "capann_dme", "capann_hsp")

periods <- tribble(
  ~file,   ~period,         ~year,
  "2014",  "2014",          2014L,
  "2015",  "2015",          2015L,
  "2016",  "2016",          2016L,
  "2017",  "2017",          2017L,
  "2018",  "2018",          2018L,
  "2019",  "2019",          2019L,
  "2019A", "2019 Jul–Dec",  2019L,
  "2020",  "2020",          2020L,
  "2021",  "2021",          2021L,
  "2022",  "2022",          2022L,
  "2023",  "2023",          2023L,
  "2024",  "2024",          2024L
)

read_puf <- function(file) {
  df <- read_csv(file.path(data_dir, paste0(file, ".csv")),
                 col_types = cols(.default = "c"), show_col_types = FALSE) |>
    clean_names() |>
    rename_with(~ str_replace_all(.x, "^cap_ann_", "capann_")) |>
    rename_with(~ str_replace_all(.x, "amb_pay$", "ambpay"))

  # Sav_rate is a fraction (2014–2020), "5.50%" text (2019A, 2021–2023),
  # or a plain percent (2024): detect per file and convert to percentage points
  sav_raw <- df$sav_rate
  sav     <- parse_number(sav_raw, na = c("", "NA", "*"))
  is_pct  <- any(grepl("%", sav_raw)) || max(abs(sav), na.rm = TRUE) > 1

  tibble(
    file     = file,
    id       = if ("aco_num" %in% names(df)) paste0("N:", df$aco_num) else paste0("ID:", df$aco_id),
    aco_name = df$aco_name,
    sav_rate = if (is_pct) sav else 100 * sav
  ) |>
    bind_cols(map_dfc(set_names(cap_vars), ~ parse_number(df[[.x]], na = c("", "NA", "*"))))
}

aco_raw <- map_dfr(periods$file, read_puf) |>
  left_join(periods, by = "file") |>
  mutate(period = factor(period, levels = periods$period))
```

```{r}
# Link 2014–2017 masked IDs to 2018+ CMS IDs by normalized ACO name
# (robustness check only; ambiguous names are left unlinked)
norm_name <- function(x) str_squish(str_replace_all(toupper(str_replace_all(x, "&", " AND ")), "[^A-Z0-9 ]", " "))

old_names <- aco_raw |>
  filter(year <= 2017) |>
  group_by(id) |> slice_max(year, n = 1, with_ties = FALSE) |> ungroup() |>
  transmute(id, key = norm_name(aco_name))

new_names <- aco_raw |>
  filter(year >= 2018) |>
  distinct(new_id = id, key = norm_name(aco_name)) |>
  group_by(key) |> filter(n_distinct(new_id) == 1) |> ungroup()

name_links <- old_names |>
  inner_join(new_names, by = "key") |>
  group_by(key) |> filter(n() == 1) |> ungroup() |>
  group_by(new_id) |> filter(n() == 1) |> ungroup() |>
  select(id, new_id)

aco_raw <- aco_raw |>
  left_join(name_links, by = "id") |>
  mutate(id_linked = coalesce(new_id, id)) |>
  select(-new_id)
```

`r nrow(name_links)` of `r n_distinct(aco_raw$id[aco_raw$year <= 2017])`
2014–2017 ACOs are linked to a 2018+ CMS ID by name.

```{r}
# Trim savings-rate outliers (1.5 x IQR) within each period
aco_data <- aco_raw |>
  group_by(period) |>
  filter(between(sav_rate,
                 quantile(sav_rate, 0.25, na.rm = TRUE) - 1.5 * IQR(sav_rate, na.rm = TRUE),
                 quantile(sav_rate, 0.75, na.rm = TRUE) + 1.5 * IQR(sav_rate, na.rm = TRUE))) |>
  ungroup()

# Spending shares as a percent of the eight categories, plus the savings indicator
aco_data <- aco_data |>
  mutate(total_exp = rowSums(across(all_of(cap_vars)), na.rm = TRUE)) |>
  mutate(across(all_of(cap_vars), ~ 100 * .x / total_exp, .names = "{.col}_pct")) |>
  mutate(saved = as.integer(sav_rate > 0))

write_csv(aco_data, file.path(data_dir, "aco_panel_2014_2024.csv"))

aco_raw |>
  count(period, name = "ACOs (raw)") |>
  left_join(
    aco_data |>
      group_by(period) |>
      summarise(`ACOs (after trimming)` = n(),
                `Median savings rate (%)` = round(median(sav_rate), 2),
                `Share that saved (%)`    = round(100 * mean(saved), 1)),
    by = "period"
  ) |>
  rename(Period = period) |>
  knitr::kable(caption = "MSSP ACOs by performance period")
```

## Spending Composition, 2014–2024

```{r}
share_labels <- c(
  capann_inp_all_pct = "Inpatient",
  capann_opd_pct     = "Outpatient (OPD)",
  capann_snf_pct     = "SNF",
  capann_pb_pct      = "Physician",
  capann_ambpay_pct  = "Ambulance",
  capann_hha_pct     = "Home Health",
  capann_dme_pct     = "DME",
  capann_hsp_pct     = "Hospice"
)

shares_by_period <- aco_data |>
  group_by(period) |>
  summarise(across(all_of(names(share_labels)), ~ mean(.x, na.rm = TRUE))) |>
  pivot_longer(-period, names_to = "category", values_to = "share") |>
  mutate(category = factor(share_labels[category], levels = share_labels))

p_shares <- ggplot(shares_by_period, aes(x = period, y = share, group = category)) +
  annotate("rect", xmin = "2020", xmax = "2021", ymin = -Inf, ymax = Inf, alpha = 0.12) +
  geom_line() +
  geom_point(size = 1.3) +
  facet_wrap(~ category, scales = "free_y", ncol = 4) +
  labs(title = "Average spending composition of MSSP ACOs, 2014–2024",
       subtitle = "Shaded: COVID-19 years (2020–2021)",
       x = NULL, y = "Share of total expenditures (%)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))

ggsave(file.path(fig_dir, "spending_shares_2014_2024.png"), p_shares, width = 10, height = 5.5, dpi = 200, bg = "white")
p_shares
```

### Year Fixed-Effects on Spending Shares

The original study's first model: each spending share regressed on year and ACO
fixed effects, so each year coefficient is the average within-ACO change since
2014 (percentage points). Shares are used rather than per-capita dollars, which
are nominal and would mostly reflect price growth over ten years. Year effects
after 2017 are only identified through ACOs observed in both ID eras, so these
models use the name-linked IDs.

```{r}
trend_data <- aco_data |> mutate(unit = id_linked)

share_trends <- map(names(share_labels), ~ feols(
  as.formula(paste(.x, "~ i(period, ref = '2014') | unit")),
  data = trend_data, cluster = ~ unit, fixef.rm = "singleton"
)) |>
  set_names(share_labels)

trend_notes <- "Outcome is each category's share of total expenditures (%). Coefficients are changes relative to 2014 (pp). ACO fixed effects (name-linked IDs); SEs clustered by ACO."

modelsummary(share_trends, coef_rename = \(x) sub("^period::", "", x), stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Year Fixed-Effects on Spending Shares, 2014–2024",
             notes = trend_notes,
             output = file.path(out_dir, "extension_spending_trends.html"))

modelsummary(share_trends, coef_rename = \(x) sub("^period::", "", x), stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Year Fixed-Effects on Spending Shares, 2014–2024",
             notes = trend_notes,
             output = "html")
```

```{r, fig.height = 5.5}
trend_coefs <- imap_dfr(share_trends, function(m, category) {
  ct <- coeftable(m)
  ci <- confint(m)
  tibble(category = category, period = sub("^period::", "", rownames(ct)),
         estimate = ct[, "Estimate"], low = ci[, 1], high = ci[, 2])
}) |>
  bind_rows(tibble(category = names(share_trends), period = "2014", estimate = 0, low = 0, high = 0)) |>
  mutate(period   = factor(period, levels = levels(aco_data$period)),
         category = factor(category, levels = share_labels))

p_trends <- ggplot(trend_coefs, aes(x = period, y = estimate, group = category)) +
  annotate("rect", xmin = "2020", xmax = "2021", ymin = -Inf, ymax = Inf, alpha = 0.12) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_ribbon(aes(ymin = low, ymax = high), alpha = 0.2) +
  geom_line() +
  geom_point(size = 1.3) +
  facet_wrap(~ category, scales = "free_y", ncol = 4) +
  labs(title = "Within-ACO change in spending shares since 2014 (pp)",
       subtitle = "Year fixed effects with ACO fixed effects, 95% CIs (ACO-clustered). Shaded: COVID-19 years",
       x = NULL, y = "Change vs. 2014 (pp)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 7))

ggsave(file.path(fig_dir, "spending_trends_2014_2024.png"), p_trends, width = 10, height = 5.5, dpi = 200, bg = "white")
p_trends
```

## Savings Models, 2014–2024

```{r}
# Same specifications as the replication; OPD share is the omitted category
fml_full    <- c("capann_inp_all_pct", "capann_hsp_pct", "capann_snf_pct", "capann_pb_pct",
                 "capann_ambpay_pct", "capann_hha_pct", "capann_dme_pct")
fml_reduced <- c("capann_inp_all_pct", "capann_hsp_pct", "capann_snf_pct", "capann_pb_pct", "capann_hha_pct")

# ACO (`unit`) and period fixed effects, SEs clustered by ACO
twfe <- function(outcome, rhs) as.formula(paste(outcome, "~", paste(rhs, collapse = " + "), "| unit + period"))

# ACOs seen in only one period add no within-ACO variation, so they are dropped
fit_fe    <- function(rhs, data) feols(twfe("sav_rate", rhs), data = data, cluster = ~ unit, fixef.rm = "singleton")
fit_logit <- function(rhs, data) feglm(twfe("saved", rhs), data = data, family = binomial(), cluster = ~ unit)

coef_labels <- c(
  capann_inp_all_pct = "Percent Inpatient",
  capann_hsp_pct     = "Percent Hospice",
  capann_snf_pct     = "Percent SNF",
  capann_pb_pct      = "Percent Physician",
  capann_ambpay_pct  = "Percent Ambulance",
  capann_hha_pct     = "Percent Home Health",
  capann_dme_pct     = "Percent DME"
)

full_sample <- aco_data |> mutate(unit = id)

main_models <- list(
  "FE: Savings Rate (Full)"    = fit_fe(fml_full, full_sample),
  "FE: Savings Rate (Reduced)" = fit_fe(fml_reduced, full_sample),
  "Logit: Full"                = fit_logit(fml_full, full_sample),
  "Logit: Reduced"             = fit_logit(fml_reduced, full_sample)
)

main_notes <- "ACO and performance-period fixed effects; SEs clustered by ACO. FE outcome is the savings rate (percentage points). Logit outcome is 1 if the ACO generated savings; coefficients are log-odds. Outpatient (OPD) share is the omitted category. The FE logit drops ACOs whose outcome never varies."

modelsummary(main_models, coef_map = coef_labels, stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Spending Composition and ACO Savings, 2014–2024",
             notes = main_notes,
             output = file.path(out_dir, "extension_main_results.html"))

modelsummary(main_models, coef_map = coef_labels, stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Spending Composition and ACO Savings, 2014–2024",
             notes = main_notes,
             output = "html")
```

## Robustness

Each column re-estimates the full specification on a different sample:

- **2014–2017**: the replication window, now with two-way fixed effects
- **2014–2024**: all periods (main estimate)
- **Excl. 2020–21**: drops the COVID-19 years
- **2014–2018 / 2019–2024**: before and after the 2019 *Pathways to Success* redesign
  (2014–2018 uses name-linked IDs so 2018 records connect to the same ACOs' earlier years)
- **Name-linked IDs**: all periods, with 2014–2017 ACOs linked to their 2018+ CMS IDs

```{r}
samples <- list(
  "2014–2017"        = aco_data |> filter(year <= 2017)              |> mutate(unit = id),
  "2014–2024"        = aco_data                                      |> mutate(unit = id),
  "Excl. 2020–21"    = aco_data |> filter(!year %in% 2020:2021)      |> mutate(unit = id),
  "2014–2018"        = aco_data |> filter(year <= 2018)              |> mutate(unit = id_linked),
  "2019–2024"        = aco_data |> filter(year >= 2019)              |> mutate(unit = id),
  "Name-linked IDs"  = aco_data                                      |> mutate(unit = id_linked)
)

robust_fe    <- map(samples, ~ fit_fe(fml_full, .x))
robust_logit <- map(samples, ~ fit_logit(fml_full, .x))

modelsummary(robust_fe, coef_map = coef_labels, stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Robustness: Fixed-Effects Regression on Savings Rate (Full Specification)",
             notes = "ACO and performance-period fixed effects; SEs clustered by ACO.",
             output = "html")

modelsummary(robust_logit, coef_map = coef_labels, stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Robustness: Fixed-Effects Logit on Achieving Savings (Full Specification)",
             notes = "ACO and performance-period fixed effects; SEs clustered by ACO. Coefficients are log-odds.",
             output = "html")

modelsummary(c(setNames(robust_fe, paste("FE:", names(robust_fe))),
               setNames(robust_logit, paste("Logit:", names(robust_logit)))),
             coef_map = coef_labels, stars = TRUE,
             estimate = "{estimate}{stars}", statistic = "({std.error})",
             gof_omit = "IC|Log|Adj|RMSE|Std.Errors",
             title = "Robustness: FE Savings-Rate and FE Logit Models Across Samples",
             notes = "ACO and performance-period fixed effects; SEs clustered by ACO.",
             output = file.path(out_dir, "extension_robustness.html"))
```

```{r, fig.height = 5.5}
key_terms <- c("capann_inp_all_pct", "capann_snf_pct", "capann_pb_pct", "capann_hha_pct")

stability <- imap_dfr(robust_fe, function(m, sample) {
  ct <- coeftable(m)
  ci <- confint(m)
  tibble(sample = sample, term = rownames(ct), estimate = ct[, "Estimate"],
         low = ci[, 1], high = ci[, 2])
}) |>
  filter(term %in% key_terms) |>
  mutate(term   = factor(coef_labels[term], levels = coef_labels[key_terms]),
         sample = factor(sample, levels = names(samples)))

p_stability <- ggplot(stability, aes(x = sample, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_pointrange(aes(ymin = low, ymax = high), size = 0.3) +
  facet_wrap(~ term, ncol = 4) +
  labs(title = "Effect of a 1 pp higher spending share on the savings rate (pp)",
       subtitle = "Two-way FE estimates with 95% CIs (ACO-clustered), by sample",
       x = NULL, y = "Coefficient") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(fig_dir, "coefficient_stability.png"), p_stability, width = 10, height = 5, dpi = 200, bg = "white")
p_stability
```
