---
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: false
    fig_caption: true
fontsize: 11pt
geometry: margin=1in
---

# Medicare ACO Spending Patterns: Replication and Extension 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 220,
                      fig.width = 7, fig.height = 4.2)

suppressPackageStartupMessages({
  library(dplyr); library(readr); library(stringr); library(purrr); library(tidyr)
  library(janitor); library(ggplot2); library(broom)
  library(plm); library(lmtest); library(sandwich)
  library(fixest); library(modelsummary)
})

# Paths are relative to this file (analysis/), so knitting works from a fresh clone
data_dir <- file.path("..", "data")
out_dir  <- file.path("..", "outputs")

# outputs dir for saved tables/figures
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
```
# Executive Summary

## Question: Do ACOs that achieve shared savings reallocate spending away from inpatient and post-acute institutional care toward physician/outpatient services?

## Answer: Largely yes. In 2014–2017 MSSP data, higher inpatient and SNF shares are associated with lower savings rates in both the full and reduced fixed-effects models; physician share is small, positive and not significant. In the fixed-effects logit, only the inpatient share significantly predicts whether an ACO saves. The direction of effects mirrors the published study.

## Data: Medicare Shared Savings Program (MSSP) PUFs, 2014–2017. Unit = ACO-year.

## Methods: Fixed-effects (ACO & year) linear model of savings rate and fixed-effects logit for achieving savings; ACO-clustered SEs.

```{r}
# Prefer a ready harmonized object; fall back to CSVs if needed
if (file.exists(file.path(data_dir, "aco_data.rds"))) {
  aco_data <- readRDS(file.path(data_dir, "aco_data.rds"))
} else if (file.exists(file.path(data_dir, "aco_data.csv"))) {
  aco_data <- read_csv(file.path(data_dir, "aco_data.csv"), show_col_types = FALSE) |> clean_names()
} else {
  # Fallback: rebuild from four yearly CSVs in data/
  files <- list.files(data_dir, pattern = "^(2014|2015|2016|2017).*\\.csv$", full.names = TRUE)
  stopifnot(length(files) >= 4)
  years <- str_extract(basename(files), "^(2014|2015|2016|2017)")
  names(files) <- years

  read_one <- function(path, yr) {
    read_csv(path, show_col_types = FALSE) |>
      clean_names() |>
      mutate(year = as.integer(yr)) |>
      # normalize CapAnn variations
      rename_with(~ str_replace_all(.x, "^cap_ann_", "capann_")) |>
      rename_with(~ str_replace(.x, "amb_pay$", "ambpay"))
  }

  raw <- map2(files, names(files), read_one)

  needed <- c("aco_num","year","sav_rate",
              "capann_inp_all","capann_opd","capann_snf","capann_pb",
              "capann_ambpay","capann_hha","capann_dme","capann_hsp")

  safe_num <- function(x) readr::parse_number(as.character(x), na = c("", "NA", "*"))

  harmonize <- function(df) {
    miss <- setdiff(needed, names(df)); if (length(miss)) df[miss] <- NA
    df |>
      mutate(aco_num = as.character(aco_num)) |>
      mutate(across(all_of(setdiff(needed, c("aco_num","year"))), safe_num)) |>
      select(any_of(needed))
  }

  datasets2 <- map(raw, harmonize)
  aco_data <- bind_rows(datasets2) |>
    arrange(aco_num, year)
}

# sanity
stopifnot(all(c("aco_num","year","sav_rate") %in% names(aco_data)))
```


```{r}
exp_cols <- c("capann_inp_all","capann_opd","capann_snf","capann_pb",
              "capann_ambpay","capann_hha","capann_dme","capann_hsp")

# Remove savings-rate outliers (1.5 x IQR), as in ACO_Final.rmd
lower_bound <- quantile(aco_data$sav_rate, 0.25, na.rm = TRUE) - 1.5 * IQR(aco_data$sav_rate, na.rm = TRUE)
upper_bound <- quantile(aco_data$sav_rate, 0.75, na.rm = TRUE) + 1.5 * IQR(aco_data$sav_rate, na.rm = TRUE)

aco_data <- aco_data |>
  filter(sav_rate >= lower_bound, sav_rate <= upper_bound)

aco_data <- aco_data |>
  mutate(total_exp = rowSums(across(all_of(exp_cols)), na.rm = TRUE)) |>
  mutate(across(all_of(exp_cols),
                ~ 100 * (.x / if_else(total_exp == 0, NA_real_, total_exp)),
                .names = "{.col}_pct")) |>
  mutate(sav_rate = if_else(abs(sav_rate) <= 1, 100 * sav_rate, sav_rate)) |>
  mutate(saved = as.integer(sav_rate > 0))

# Panel objects
fe_data  <- pdata.frame(aco_data, index = c("aco_num","year"))
glm_data <- as.data.frame(aco_data) |>
  mutate(year = factor(year), aco_num = factor(aco_num))
```

```{r}
# Simple year-over-year shifts in spending shares (means)
desc <- aco_data |>
  summarize(across(ends_with("_pct"), ~ mean(.x, na.rm = TRUE)), .by = year) |>
  pivot_longer(-year, names_to = "var", values_to = "pct") |>
  mutate(var = recode(var,
    capann_inp_all_pct = "Inpatient",
    capann_snf_pct     = "SNF",
    capann_pb_pct      = "Physician",
    capann_opd_pct     = "Outpatient (OPD)",
    capann_hha_pct     = "Home Health",
    capann_hsp_pct     = "Hospice",
    capann_ambpay_pct  = "Ambulance",
    capann_dme_pct     = "DME"
  ))

ggplot(desc, aes(x = year, y = pct, group = var)) +
  geom_line() + geom_point() +
  labs(title = "Average Spending Composition by Year",
       x = "Year", y = "Share of Total Expenditures (%)", color = "") +
  theme_minimal(base_size = 11) +
  guides(color = "none") +
  facet_wrap(~ var, scales = "free_y")
```


```{r}
# Full & reduced FE models
fe_full <- plm(
  sav_rate ~ capann_inp_all_pct + capann_hsp_pct + capann_snf_pct +
             capann_pb_pct + capann_ambpay_pct + capann_hha_pct + capann_dme_pct,
  data = fe_data, model = "within", effect = "twoways"
)
fe_reduced <- plm(
  sav_rate ~ capann_inp_all_pct + capann_hsp_pct + capann_snf_pct +
             capann_pb_pct + capann_hha_pct,
  data = fe_data, model = "within", effect = "twoways"
)

se_fe_full    <- vcovHC(fe_full,    type = "HC1", cluster = "group")
se_fe_reduced <- vcovHC(fe_reduced, type = "HC1", cluster = "group")

labels <- c(
  capann_inp_all_pct = "Percent Inpatient",
  capann_snf_pct     = "Percent SNF",
  capann_pb_pct      = "Percent Physician",
  capann_hha_pct     = "Percent Home Health",
  capann_hsp_pct     = "Percent Hospice",
  capann_ambpay_pct  = "Percent Ambulance",
  capann_dme_pct     = "Percent DME"
)

msummary(
  list("FE: Savings Rate (Full)" = fe_full,
       "FE: Savings Rate (Reduced)" = fe_reduced),
  vcov      = list(se_fe_full, se_fe_reduced),
  coef_map  = labels,
  estimate  = "{estimate}{stars}",
  statistic = "({std.error})",
  stars     = TRUE,
  gof_omit  = "IC|Log|Adj|Within|RMSE",
  title     = "Effect of Spending Composition on ACO Savings Rate",
  notes     = "Outcome is savings rate (percentage points). ACO & year fixed effects; SEs clustered by ACO.",
  output    = file.path(out_dir, "fe_models.html")
)
```


```{r}
tidy_fe <- tidy(fe_reduced, conf.int = TRUE, vcov = se_fe_reduced) |>
  filter(term %in% names(labels)) |>
  mutate(term = recode(term, !!!labels))

ggplot(tidy_fe, aes(x = reorder(term, estimate), y = estimate)) +
  geom_point(size = 2.6) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(title = "Impact on Savings Rate (pp) per 1 pp change in spending share",
       x = "", y = "Coefficient (95% CI)") +
  theme_minimal(base_size = 11)
```


```{r}
logit_full <- feglm(
  saved ~ capann_inp_all_pct + capann_hsp_pct + capann_snf_pct +
           capann_pb_pct + capann_ambpay_pct + capann_hha_pct + capann_dme_pct |
           aco_num + year,
  data = glm_data, family = binomial()
)

logit_reduced <- feglm(
  saved ~ capann_inp_all_pct + capann_hsp_pct + capann_snf_pct +
           capann_pb_pct + capann_hha_pct |
           aco_num + year,
  data = glm_data, family = binomial()
)

msummary(
  list("Logit: Full" = logit_full,
       "Logit: Reduced" = logit_reduced),
  vcov      = ~ aco_num,   # cluster by ACO
  coef_map  = labels,
  estimate  = "{estimate}{stars}",
  statistic = "({std.error})",
  stars     = TRUE,
  gof_omit  = "IC|Log|Adj|Within",
  title     = "Probability of Achieving Savings: Fixed-Effects Logits",
  notes     = "Logit coefficients are log-odds. ACO & year fixed effects; SEs clustered by ACO.",
  output    = file.path(out_dir, "logit_models.html")
)
```


```{r}
or_tbl <- tidy(logit_reduced, conf.int = TRUE, cluster = "aco_num") |>
  filter(term %in% names(labels)) |>
  mutate(term = recode(term, !!!labels),
         OR = exp(estimate), LCL = exp(conf.low), UCL = exp(conf.high)) |>
  select(Predictor = term, `Odds ratio` = OR, `95% LCL` = LCL, `95% UCL` = UCL)

knitr::kable(or_tbl, digits = 3, caption = "Odds of achieving savings per 1 pp change in spending share")

ggplot(or_tbl, aes(x = reorder(Predictor, `Odds ratio`), y = `Odds ratio`)) +
  geom_point(size = 2.6) +
  geom_errorbar(aes(ymin = `95% LCL`, ymax = `95% UCL`), width = 0.15) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  coord_flip() +
  labs(title = "Odds Ratios from Reduced Logit",
       x = "", y = "Odds ratio (95% CI)") +
  theme_minimal(base_size = 11)
```


# Results & Interpretation

## The fixed-effects regressions indicate that ACOs achieving higher shared savings allocate less of total spending to inpatient and skilled nursing facility (SNF) care. A 1 pp higher inpatient share is associated with a roughly 0.5 pp lower savings rate, and a 1 pp higher SNF share with a roughly 0.3 pp lower rate, in both the full and reduced models. Home health shows a similar negative association (about 0.3 pp). Physician spending is small, positive and not significant. The fixed-effects logits are weaker: a higher inpatient share significantly reduces the odds of achieving savings, while the SNF, home health and physician coefficients are not significant.

## Practically, programs that prevent avoidable inpatient admissions and reduce SNF use are most closely associated with MSSP savings. The direction of these effects tracks the published findings over a later time window, suggesting the pattern is persistent.


